```{r chapter-5-setup, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE}

library(data.table)
library(tidyverse)
library(hrbrthemes)
library(ggcorrplot)
library(GoodmanKruskal)
library(lindia)
library(finalfit)
library(sigr)
library(papaja)
library(here)
library(kableExtra)
library(GGally)
library(gtools)
library(MASS)
library(corrplot)
library(magrittr)
library(caret)
library(factoextra)
library(broom)
library(ggord)
library(skimr)
library(sjstats)
library(explor)
library(stargazer)
# For overriding MASS select() 
select <- dplyr::select
set.seed(2018)
```


# Week 5 – Dimensionality reduction techniques

This week we'll try out dimensionality reduction techniques, namely principal components analysis (PCA) and multiple correspondence analysis (MCA) which can be used to explore multidimensional data as points in a low-dimensional space. PCA is applicable for continous data and MCA can be used when the measurement level of variables is at least categorical.

## Exploring human development data with principal component analysis

The dataset we'll use is  from the United Nations Development Programme. For the following analysis, the dataset was preprocessed using an [external script](https://github.com/mjlassila/IODS-project/blob/mjlassila/chapter-5/data/create_human.R).

The dataset has the following variables:

   Code                Description
  -----------------   ------------------------------
   country            name of the country (as row name)
   edu_exp            expected years of education
   sec_edu_female     proportion of females at least 2nd level education
   labour_rate_female proportion of women participating in working life
   life_exp           life expectancy at birth
   gni                gross national income
   mat_mortality      maternal mortality ratio
   birth_rate         the number of births to women ages 15–19 (per 1000 women)
   repr_parlament     percetange of female members of the parliament



```{r chapter-5-load-data, echo = FALSE, message = FALSE, cache = TRUE}
human <- read.csv(here("data/human.csv"), row.names = 1)
```


```{r eval = FALSE, echo = FALSE}
# Show a graphical overview of the data and show summaries of the variables in
# the data. Describe and interpret the outputs, commenting on the distributions
# of the variables and the relationships between them. (0-2 points)
```

```{r chapter-5-graphical-overview, fig.width = 9, fig.height = 9, echo = FALSE, cache = TRUE}
GGally::ggpairs(human)
```

Most of the variable distributions are more-or-less skewed, so it is very likely that PCA will not work without normalizing the data. Only the edu_exp (expected years of education) variable looks normally distributed. 

```{r chapter-5-summary-table, echo = FALSE, results = "asis" }
stargazer(
    human, 
    type = "html", 
    summary.stat = c("min", "p25", "median", "p75", "max", "median", "sd"),
    digit.separator = "",
    digits = 2,
    decimal.mark = ','
)
```



```{r human-corrplot, eval = TRUE, echo = FALSE, fig.width = 9, fig.height = 9}

corrplot(
  cor(human),
  method = "number"
)
```

Unexpectedly, maternal mortality and life expectency are highly correlated (-0.86) as well as high birth rate and maternal mortality (0.76). Variables regarding education have also strong correlation with maternal mortality, so that higher values in education variables, lead to lower value in maternal mortality rate. Secondary education of women have an association to lower birth rate (-0.68), as well as higher life expectancy (-0.72). Quite surprisingly, women's representation in the parliament doesn't seem to have an association with any of other variables or the association is very low. As one might expect, maternal mortality and high birth rate are associated with low GNI.

```{r eval = FALSE, echo = FALSE}
# Perform principal component analysis (PCA) on the not standardized human data.
# Show the variability captured by the principal components. Draw a biplot
# displaying the observations by the first two principal components (PC1
# coordinate in x-axis, PC2 coordinate in y-axis), along with arrows
# representing the original variables. (0-2 points)
```

#### First round of PCA

```{r pca-incorrectly}
pca_summary_rows <- c(
  "Standard deviation",
  "Eigenvalue",
  "Proportion of Variance",
  "Cumulative Proportion")

pca_human <- prcomp(human)
pca_summary <- pca(pca_human)
rownames(pca_summary) <- pca_summary_rows
```

```{r, echo = FALSE}
knitr::kable(
round(pca_summary,1),
caption = "Principal component analysis of non-standardized data"
)
```


As expected, PCA with non-standardized data doesn't make much sense as it'll happen that the variable will largest variance explains virtually all of the variance and therefore creating a biplot will fail miserably. 

```{r, echo = FALSE, fig.width=4, fig.height=1}
ggord(pca_human, arrow = 0.4, txt = 5, vec_ext = 4, size = 0.06) + theme_minimal()
```



#### Second round of PCA
```{r eval = FALSE, echo = FALSE}
# Standardize the variables in the human data and repeat the above analysis.
# Interpret the results of both analysis (with and without standardizing). Are
# the results different? Why or why not? Include captions (brief descriptions)
# in your plots where you describe the results by using not just your variable
# names, but the actual phenomenons they relate to. (0-4 points)
```

```{r pca-done-right}
# Use built-in scaling in prcomp() -function
pca_human_final <- prcomp(human, scale = TRUE, center = TRUE)
pca_summary_final <- pca(pca_human_final)
rownames(pca_summary_final) <- pca_summary_rows
```


```{r, echo = FALSE}
knitr::kable(
round(pca_summary_final,2),
caption = "Principal component analysis of standardized data"
)
```
After scaling, PCA results will make sense. Because in the original data, variance and distribution vary wildly, using PCA in non-standardized data will yield strange results as PCA as a analysis technique is sensitive to to the relative scaling of variables.

```{r, echo = FALSE, fig.width=10, fig.height=9, cache = FALSE}
ggord(
  pca_human_final,
  arrow = 0.4,
  txt = 3,
  vec_ext = 4.5,
  size = 2,
  obslab = TRUE,
  repel = TRUE,
  labcol = "red") + theme_minimal()
```

```{r eval = FALSE, echo = FALSE}
# Give your personal interpretations of the first two principal component
# dimensions based on the biplot drawn after PCA on the standardized human data.
# (0-2 points)
```

The first principal component explains 57.03% of the total variance of the dataset, and the second principal component explains 15.6% of it. It seems that the first principal component represents *classical welfare state*, eg. healthcare and access to education. The second principal component represents *gender equality*, eg. do women have equal rights to participate in politics and working life. Looking the biplot, it is evident that having an all-embracing welfare state doesn't necessarily make a country gender-equal and vice-versa, Qatar (and other oil-rich states) and Rwanda exemplifying this phenomenon.

Our initial exploration of the data using a correlation matrix revealed the complex relationship between variables partially but usage of PCA biplot makes it possible to understand the whole dataset in an intuitive way almost at a glance.

## Exploring tea survey data with multiple correspondence analysis

```{r eval = FALSE, echo = FALSE}
# Load the tea dataset from the package Factominer. Explore the data briefly:
# look at the structure and the dimensions of the data and visualize it.

```
```{r, echo = FALSE}
library(FactoMineR)
data(tea)
```



The following questions which were used to gather the tea dataset can be found from the book [Exploratory Multivariate Analysis by Example Using R](https://books.google.fi/books/about/Exploratory_Multivariate_Analysis_by_Exa.html?id=nLrODgAAQBAJ&redir_esc=y), Second Edition, pages 131–132.

1. What kind of tea do you drink the most (black tea, green tea, flavored tea)
2. How do you take your tea (nothing added, with lemon, with milk, other)
3. What kind of tea you buy ? (tea bags, loose tea, both)
4. Do you add sugar to your tea (yes, no)?
5. Where do you buy your tea (supermarket, specialist shop, both)
6. What kind of tea you buy (cheapest supermarket brand, well-known brand, upscale, it varies)
7. How ofter do you drink tea (more than twice a day, once a day, 3 to 6 times a week, once or twice per week)?

Location in wich the tea is drunk (yes/no):

8. Do you drink tea at home?
9. Do you drink tea at work?
10. Do you drink tea in tearooms or coffee shops?
11. Do you drink tea at frinds houses?
12. Do you drink tea in restaurants?
13. Do you drink tea in bars?

Time of the day (yes/no): 

14. Do you drink tea at breakfast?
15. Do you drink tea in the afternoon?
16. Do you drink tea in the evening?
17. Do you drink tea after lunch?
18. Do you drink tea after dinner?
19. Do you drink tea throughout the day?

Questions about the image of the product (yes/no):

20 Do you consider tea to be exotic?
21. Do you associate tea with spirituality?
22. Is tea good for your health?
23. Is tea a diuretic?
24. Do you associate tea with friendliness?
25. Does tea stop the body from absorbing iron?
26. Is tea feminine?
28. Is tea refined?
29. Will tea help you to lose weight?
30. Is tea a stimulant?
31. Is tea a relaxant?
32. Does tea have any effect on your overall health?

Additionally, there were four background questions regarding:

* sex
* professional category (farmer, manual labourer, professional, senior management, employee, other profession, student, unemployed)
* age
* regular participation in sports (yes or no).

```{r variable-overview, results='asis', echo = FALSE}
skim(tea) %>% skimr::kable()
```

```{r chapter5-explore-data-overview, cache = TRUE, echo = FALSE}

taumatrix <- GKtauDataframe(tea %>% select(-age))
```
Let's use [Goodman-Kruskal tau](https://cran.r-project.org/web/packages/GoodmanKruskal/vignettes/GoodmanKruskal.html) to explore associations between variables in the dataset. Goodman-Kruskal tau can be used to study association when the measurement level of the variable is at least categorical.



```{r chapter-5-tau-matrix, fig.width=17, fig.height=14, echo = FALSE, cache = TRUE}
plot(taumatrix,colorPlot = FALSE,diagSize = 0.15)
```

Going trough the Goodman-Kruskal tau-matrix, we can observe few interesting associations between variables. Overall, associations are not very strong, but the following associations emerge when taking into account associations which are stronger than the mean tau-value (0.08).

* gender and considering tea feminine or not feminine
* age and considering tea feminine or not feminine
* age and a belief that tea has an effect on overall health
* a belief that tea has an effect on overall health and and considering tea healthy
* type of the tea and age
* type of the tea and habit of using sugar
* drinking tea at tearoom and buying tea from specialist shops
* frequency of drinking tea and drinking tea at breakfast
* price, type of tea and the place of purchase

Even though we'll eventually use all variables in the MCA, let's observe few of these associations bit more closely, just for the sake of curiosity.

```{r chapter-5-tea-pair-comparison, cache = FALSE, echo = FALSE, fig.width=13, fig.height=13}
ggpairs(tea %>% 
  select(sex,age_Q,feminine,Tea,where,price))
```

* There are more female than male responders to the survey. Over the half of the responders are under 35.
* It seems that the bulk of the tea is being bought from the chain stores and the preferred style is Earl Gray.
* Interestingly, considering all answers tea is not considered feminine, but females tend to associate tea with femininity more often than men.

### Drinkers of different types of tea 

Let's use MCA to investigate if there are observable differences between groups of people who prefer different types of tea.

```{r, eval = FALSE, echo = FALSE}
# Then do Multiple Correspondence Analysis on the tea data (or to a certain
# columns of the data, it’s up to you). Interpret the results of the MCA and
# draw at least the variable biplot of the analysis.
```
```{r mca-with-factominer}
tea_mca <- MCA(tea,
    quanti.sup = 19, # age as quantitative supplement
    quali.sup = c(20:36),
    graph = FALSE) # other background variables as categorical 
                                # supplements
```



```{r variable-biplot, fig.width=10, fig.height=9}
ggord(tea_mca,
      tea$Tea,
      arrow = 0.2,
      txt = 3,
      vec_ext = 1.05,
      size = 2,
      repel = TRUE) + theme_minimal()
```

Looking the biplot graph, drinkers of different types of tea seem to somewhat differ. People who drink green and black tea are more often buying their tea from specialist shops, prefer unpackaged tea and dont't mind spending more money. Drinking Earl Gray seems to serve similiar purpose as habitual coffee drinking in Finland – people drink it at work or with friends during lunch.

It should be kept in mind that the proportion of green tea drinkers is very small compared to the friends of Earl Gray so that these conclusions based on biplot are likely not very strong.



```{r, eval = FALSE, echo = FALSE}
# You can also explore other plotting options for MCA. Comment on the output of
# the plots. (0-4 points)
```

#### Extra plots

The following interactive plot makes it possible to investigate which responses have the strongest contribution to the model. Darker the colour, stronger the contribution.

```{r contributions-plot, eval = TRUE, echo = FALSE }
res <- explor::prepare_results(tea_mca)
explor::MCA_ind_plot(res, xax = 1, yax = 2,ind_sup = FALSE,
    lab_var = NULL, , ind_lab_min_contrib = 0,
    col_var = NULL, labels_size = 9,
    point_opacity = 0.5, opacity_var = "Contrib", point_size = 54,
    ellipses = FALSE, transitions = TRUE, labels_positions = NULL,
    xlim = c(-1, 1.44), ylim = c(-1.06, 1.38))
```


*Scree plot* cat be used to visualize the percentage of inertia (eg. in this context, variance) explained by MCA dimensions. From the plot we can see that the first dimension explains approximately 10% percent of variance and the second one 8%. 

```{r eigvenvalues, cache = TRUE, eval = TRUE, echo = FALSE}
fviz_screeplot(tea_mca,geom = "bar")
```







