```{r chapter-3-setup}
library(here)
library(PerformanceAnalytics)
library(data.table)
library(tidyverse)
library(hrbrthemes)
library(ggcorrplot)
library(GoodmanKruskal)
library(lindia)
library(finalfit)
library(sigr)
library(magrittr)
library(ggmosaic)
```

## Week 3 – Logistic regression

```{r load-data}

# Print out the names of the variables in the data and describe the data set
# briefly, assuming the reader has no previous knowledge of it. 
# There is information related to the data here. (0-1 point)

alc <- fread(here("data/chapter3_alc.csv"))
names(alc)
# skim(alc)
```

The original dataset has the following variables:


* school - student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira) 
* sex - student's sex (binary: 'F' - female or 'M' - male) 
* age - student's age (numeric: from 15 to 22) 
* address - student's home address type (binary: 'U' - urban or 'R' - rural) 
* famsize - family size (binary: 'LE3' - less or equal to 3 or 'GT3' - greater than 3) 
* pstatus - parent's cohabitation status (binary: 'T' - living together or 'A' - apart) 
* medu - mother's education (numeric: 0 - none, 1 - primary education (4th grade), 2  - 5th to 9th grade, 3 - secondary education or 4 - higher education) 
* fedu - father's education (numeric: 0 - none, 1 - primary education (4th grade), 2 - 5th to 9th grade, 3 - secondary education or 4 - higher education) 
* mjob - mother's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other') 
* fjob - father's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other') 
* reason - reason to choose this school (nominal: close to 'home', school 'reputation', 'course' preference or 'other') 
* guardian - student's guardian (nominal: 'mother', 'father' or 'other') 
* traveltime - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour) 
* studytime - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours) 
* failures - number of past class failures (numeric: n if 1<=n<3, else 4) 
* schoolsup - extra educational support (binary: yes or no) 
* famsup - family educational support (binary: yes or no) 
* paid - extra paid classes within the course subject (Math or Portuguese) (binary: yes or no) 
* activities - extra-curricular activities (binary: yes or no) 
* nursery - attended nursery school (binary: yes or no) 
* higher - wants to take higher education (binary: yes or no) 
* internet - Internet access at home (binary: yes or no) 
* romantic - with a romantic relationship (binary: yes or no) 
* famrel - quality of family relationships (numeric: from 1 - very bad to 5 - excellent) 
* freetime - free time after school (numeric: from 1 - very low to 5 - very high) 
* goout - going out with friends (numeric: from 1 - very low to 5 - very high) 
* dalc - workday alcohol consumption (numeric: from 1 - very low to 5 - very high) 
* walc - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high) 
* health - current health status (numeric: from 1 - very bad to 5 - very good) 
* absences - number of school absences (numeric: from 0 to 93) * 
* g1 - first period grade (numeric: from 0 to 20) 
* g2 - second period grade (numeric: from 0 to 20) 
* g3 - final grade (numeric: from 0 to 20, output target)

In addition, the following variables were created for the analysis

* alc_use - mean of workday and weekend alcohol consumption (numeric)
* high_use - binary variable, TRUE if alc_use > 2


```{r prepare-dataset}
categorical_colums <- c(
  "school",
  "sex",
  "address",
  "famsize",
  "pstatus",
  "medu",
  "fedu",
  "mjob",
  "fjob",
  "reason",
  "guardian",
  "traveltime",
  "studytime",
  "failures",
  "schoolsup",
  "famsup",
  "paid",
  "activities",
  "nursery",
  "higher",
  "internet",
  "romantic",
  "famrel",
  "freetime",
  "goout",
  "high_use"
)

alc %<>%
       mutate_at(categorical_colums, funs(factor(.))) %>% select(categorical_colums,absences)

alc %<>% mutate(absnt = GroupNumeric(absences, n = 5)) %>% select(categorical_colums,absnt)
```



```{r explore-data-overview}
# The purpose of your analysis is to study the relationships between 
# high/low alcohol consumption and some of the other variables in the data. 
# To do this, choose 4 interesting variables in the data and for each of them
# present your personal hypothesis about their relationships with alcohol 
# consumption. (0-1 point)
taumatrix <- GKtauDataframe(alc)
```

```{r tau-matrix, fig.width=10, fig.height=8, echo = FALSE}
plot(taumatrix,colorPlot = FALSE,diagSize = 0.2)
```

```{r create-final-dataset, echo = FALSE}
alc %<>%  select(high_use,goout,studytime,absnt)

# Factor variables need to be named in meaninful way following the
# original data descroption

# studytime - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours) 
# goout - going out with friends (numeric: from 1 - very low to 5 - very high)

# abseentism
# 0-1, 2, 3-4, 5-7, 8-

# alcohol usage

# High orninary

studytime_labels <- c("<2 h","2–5 h","5–10 h",">10 h")
absnt_labels <- c("0–1", "2", "3–4", "5–7", ">8")
goout_labels <- c("vlow", "low", "mod", "high", "vhigh")
alc_labels <- c("low","high")
alc %<>% transmute(
  high_use = factor(high_use, labels = alc_labels),
  studytime = factor(studytime, labels = studytime_labels),
  absnt = factor(absnt, labels = absnt_labels),
  goout = factor(goout, labels = goout_labels)
  )



```

```{r tau-matrix-final}
taumatrix <- GKtauDataframe(alc)
plot(taumatrix,colorPlot = TRUE,diagSize = 0.6)
```


```{r visualize-final-dataset,fig.width=10, fig.height=4, echo = FALSE}
ggplot(alc) +
    geom_mosaic(aes(x=product(absnt, goout), fill=high_use)) +
    labs(
      x = "alcohol usage vs. frequency of going out",
      y = "frequency of absenteeism",
      fill='level of alcohol usage',
      title = "Relationship of partying, alcohol use and absenteeism"
    )
      
  

```

```{r visualize-final-dataset-studytime,fig.width=10, fig.height=4, echo = FALSE}
ggplot(alc) +
    geom_mosaic(aes(x=product(studytime, goout), fill=high_use)) +
    labs(
      x = "alcohol usage vs. frequency of going out",
      y = "study time",
      fill='level of alcohol usage',
      title = "Relationship of partying, alcohol use and time dedicated to studies"
    )
      
  

```

Numerically and graphically explore the distributions of your chosen variables and their relationships with alcohol consumption (use for example cross-tabulations, bar plots and box plots). Comment on your findings and compare the results of your exploration to your previously stated hypotheses. (0-5 points)

Use logistic regression to statistically explore the relationship between your chosen variables and the binary high/low alcohol consumption variable as the target variable. Present and interpret a summary of the fitted model. Present and interpret the coefficients of the model as odds ratios and provide confidence intervals for them. Interpret the results and compare them to your previously stated hypothesis. Hint: If your model includes factor variables see for example the first answer of this stackexchange thread on how R treats and how you should interpret these variables in the model output (or use some other resource to study this). (0-5 points)

Using the variables which, according to your logistic regression model, had a statistical relationship with high/low alcohol consumption, explore the predictive power of you model. Provide a 2x2 cross tabulation of predictions versus the actual values and optionally display a graphic visualizing both the actual values and the predictions. Compute the total proportion of inaccurately classified individuals (= the training error) and comment on all the results. Compare the performance of the model with performance achieved by some simple guessing strategy. (0-3 points)

Bonus: Perform 10-fold cross-validation on your model. Does your model have better test set performance (smaller prediction error using 10-fold cross-validation) compared to the model introduced in DataCamp (which had about 0.26 error). Could you find such a model? (0-2 points to compensate any loss of points from the above exercises)

Super-Bonus: Perform cross-validation to compare the performance of different logistic regression models (= different sets of predictors). Start with a very high number of predictors and explore the changes in the training and testing errors as you move to models with less predictors. Draw a graph displaying the trends of both training and testing errors by the number of predictors in the model. (0-4 points to compensate any loss of points from the above exercises)



